---
title: "Agriculture Analysis"
author: "Faith"
format: html
editor: visual
---

## Libraries

```{r}
library(tidyverse)
library(sf)
library(here)
library(mapview)
library(RColorBrewer)
```

## Reading in data

```{r}
ag_2015 <- st_read(here("data/ag_2015.geojson"))
ag_2020 <- st_read(here("data/ag_2020.geojson"))
```

## EDA

```{r}
mapview(ag_2015, zcol= "cropcatego")
##use colorbrewer to enhance color scale 
```

### Distribution of acreage

```{r}
ggplot(ag_2015, aes(x=acreage)) +
  geom_histogram() + 
  theme_bw() +
  facet_wrap(~island, scales="free_y")
```

## Questions: 

2015-\>2020. Did we add/lose ag land acreage between 2015 and 2020?

```{r}
sum_2015_ac <- sum(ag_2015$acreage)
sum_2020_ac <- sum(ag_2020$acreage)

sum_2015_ac - sum_2020_ac
```

We lost 27,033 acres of agriculture land.

#### What is this land now being used for?

one idea, is a map showing the loss of ag land between 2015 and 2020.

ag land ownership.

What type of agriculture/crop had the most loss?

Types of visualizations to answer these questions:

maps:

Diagrams:

Sankey diagrams
