---
title: "lights_presentation"
format: html
editor: visual
---

## **DS-422 Geocomputation: Kona Low Storms**

**Research question:** How did the first Kona Low storm and second Kona Low storm differ in their impact on power outages on Oahu?

**Sub questions:**

1.  Regionally, how did the impacts differ? 

2.  Did the outage patterns correlate with the storm's paths?

### **Part 1:**

**Baseline power outages: What did nighttime light luminance look like before the kona low storms:** 

- Week before first storm, period of 7 days: 1-7th of march.

```{r}
#| message: false
#| warning: false

library(tidyverse)
library(tigris)
library(sf)
library(blackmarbler)
library(purrr)
library(terra)
library(tidyterra)
library(lubridate)
library(here)
library(exactextractr)
library(tmap)
library(gt)

options(tigris_use_cache = TRUE)

bearer <- Sys.getenv("NASA_BEARER")
```

Cropping the ROI to Oahu only, since the map would be mostly ocean

```{r}
oahu <- counties(state = "HI", cb = TRUE, year = 2022) |>
  st_transform(4326) |>
  filter(NAME == "Honolulu") |>
  st_crop(xmin = -158.35, xmax = -157.60, ymin = 21.20, ymax = 21.75)

plot(st_geometry(oahu), main = "Oahu ROI")
```

Baseline dates:

These are the dates I want to use as a baseline comparison to my kona low storms. Because night time light luminescence can fluctuate without the influence of storms or power outages, it is necessary to have a independent variable of luminence to compare our outages too.

VNP46A2= Daily product

tryCatch()= wrapper to prevent batch from failing (since some dates are missing based on previous findings.)

```{r}
pull_night <- function(d) {
  tryCatch({
    bm_raster(roi_sf = oahu,
              product_id = "VNP46A2",
              date = d,
              bearer = bearer,
              quality_flag_rm = 2)
  }, error = function(e) {
    message("No imagery for ", d)
    NULL
  })
}

pull_period <- function(dates) {
  lst <- map(dates, pull_night)
  names(lst) <- as.character(dates)
  lst <- compact(lst)
  if (length(lst) == 0) return(NULL)
  r <- rast(lst)
  names(r) <- names(lst)          # claude said blackmarbler renames layers, so force it back
  terra::mask(r, oahu)
}
```

```{r}
baseline_dates <- seq.Date(as.Date("2026-03-01"), as.Date("2026-03-07"), by = "day")
r_baseline <- pull_period(baseline_dates)

message("Baseline nights retrieved: ", nlyr(r_baseline), " of 7")
message("Dates: ", paste(names(r_baseline), collapse = ", "))

baseline_mean <- mean(r_baseline, na.rm = TRUE)
baseline_sd   <- app(r_baseline, sd, na.rm = TRUE)
```

```{r}
storm1_dates <- seq.Date(as.Date("2026-03-10"), as.Date("2026-03-16"), by = "day")
storm2_dates <- seq.Date(as.Date("2026-03-19"), as.Date("2026-03-24"), by = "day")
gap_date     <- as.Date("2026-03-18")

r_storm1 <- pull_period(storm1_dates)
r_storm2 <- pull_period(storm2_dates)
r_gap    <- pull_period(gap_date)

storm1_mean <- mean(r_storm1, na.rm = TRUE)
storm2_mean <- mean(r_storm2, na.rm = TRUE)
gap_mean    <- mean(r_gap, na.rm = TRUE)

tibble(
  period = c("Baseline", "Storm 1", "Gap (Mar 18)", "Storm 2"),
  requested = c(7, 7, 1, 6),
  retrieved = c(nlyr(r_baseline), nlyr(r_storm1), nlyr(r_gap), nlyr(r_storm2))
) |> gt() |> tab_header(title = "Imagery Retrieved by Period")
```

```{r}
#| fig-width: 11
#| fig-height: 9

log_it <- function(r) { r[] <- log(r[] + 1); r }

panel <- c(log_it(baseline_mean), log_it(storm1_mean),
           log_it(gap_mean),      log_it(storm2_mean))

names(panel) <- c("Baseline (Mar 1-7)", "Storm 1 (Mar 10-16)",
                  "Gap (Mar 18)",       "Storm 2 (Mar 19-24)")

shared_limits <- range(values(panel), na.rm = TRUE)

ggplot() +
  geom_spatraster(data = panel) +
  facet_wrap(~lyr, ncol = 2) +
  scale_fill_gradient2(low = "black", mid = "yellow", high = "red",
                       midpoint = mean(shared_limits),
                       limits = shared_limits,
                       na.value = "transparent",
                       name = "log(radiance + 1)") +
  labs(title = "Oahu Nighttime Lights Across the March 2026 Kona Lows",
       subtitle = "All four panels share one color scale, so brightness is directly comparable") +
  coord_sf() +
  theme_void() +
  theme(plot.title = element_text(face = "bold", hjust = 0.5),
        strip.text = element_text(face = "bold"))
```

### Part 2

Storm 1 power outages: What did nighttime light luminance look like from March 10-16th (note, missing satellite imagery for a few of these days. Will include this shortcoming in conclusion.)

### Part 3:

Storm 2 power outages: What did nighttime light luminance look like from March 19-24th? 

### Part 4:

Gap coverage: What did nighttime light luminance look like on March 18th? (for a better gauge on the luminance change between storms.)

### Part 5:

Using the Hawaii Climate Data Portal to compare rainfall coverage to power outages on Oahu
