---
title: "nighttime_lights"
author: "Faith &Ashley"
format: html
editor: visual
---

## **DS-422 Geocomputation: Kona Low Storms**

**Research question:** How did the first Kona Low storm and second Kona Low storm differ in their impact on power outages on Oahu?

***Sub questions:***

1.  *Regionally, how did the impacts differ?*

2.  *Did the outage patterns correlate with the storm's paths?*

### setup

```{r}
#| message: false
#| warning: false

library(blackmarbler)
library(geodata)
library(sf)
library(terra)
library(ggplot2)
library(plotly)
library(dplyr)
library(tidyterra)
library(lubridate)
library(tigris)

options(tigris_use_cache = TRUE)

bearer <- Sys.getenv("NASA_BEARER")
```

### Dates

Baseline dates: March 1 to march 7

Storm1: March 10 to March 16

Gap: March 18th

storm 2: March 19-24

```{r}
all_dates <- list(
  baseline = seq.Date(from = ymd("2026-03-01"), to = ymd("2026-03-07"), by = "day"),
  gap      = ymd("2026-03-18"),
  storm1   = seq.Date(from = ymd("2026-03-10"), to = ymd("2026-03-16"), by = "day"),
  storm2   = seq.Date(from = ymd("2026-03-19"), to = ymd("2026-03-24"), by = "day")
)
```

Hawaii boundary

```{r}
roi_hawaii <- counties(state = "HI", cb = TRUE, year = 2022) |>
  st_transform(4326) |>
  st_crop(xmin = -160.5, xmax = -154.5, ymin = 18.5, ymax = 22.5) |>
  filter(NAMELSAD != "Kalawao County") |>
  mutate(county = gsub(" County", "", NAME))
```

```{r}
#| message: false
#| warning: false
r_list <- list()
for (period in names(all_dates)) {
  r_list[[period]] <- bm_raster(roi_sf = roi_sf,
                                product_id = "VNP46A2",  
                                quality_flag_rm = 2,
                                date = all_dates[[period]],
                                bearer = bearer)
}
```

```{r}
r_masked <- list()
for (period in names(r_list)) {
  r <- r_list[[period]] |> terra::mask(vect(roi_sf))
  r[] <- log(r[] + 1)
  r_masked[[period]] <- r
}
```

```{r}
period_mean <- list()
for (period in names(r_masked)) {
  period_mean[[period]] <- terra::mean(r_masked[[period]], na.rm = TRUE)
}

```

```{r}
### Mean raster per period + change rasters

period_mean <- list()
for (period in names(r_masked)) {
  period_mean[[period]] <- terra::mean(r_masked[[period]], na.rm = TRUE)
}

absolute_storm1 <- period_mean$storm1 - period_mean$baseline
absolute_storm2 <- period_mean$storm2 - period_mean$baseline
absolute_storm2_gap <- period_mean$storm2 - period_mean$gap

percent_storm1 <- ((period_mean$storm1 - period_mean$baseline) / period_mean$baseline) * 100
percent_storm2 <- ((period_mean$storm2 - period_mean$baseline) / period_mean$baseline) * 100
percent_storm2_gap <- ((period_mean$storm2 - period_mean$gap) / period_mean$gap) * 100

baseline_sd <- terra::app(r_masked$baseline, sd, na.rm = TRUE)
z_score_storm1 <- (period_mean$storm1 - period_mean$baseline) / baseline_sd
z_score_storm2 <- (period_mean$storm2 - period_mean$baseline) / baseline_sd
```

```{r}

### Map: baseline period, before Kona low
ggplot() +
  
  geom_spatraster(data = r_masked$baseline) +
  facet_wrap(~lyr, ncol = 4) +
  scale_fill_gradient2(low = "black",
                       mid = "yellow",
                       high = "red",
                       midpoint = 4.5,
                       na.value = "transparent") +
  
  labs(title = "Oahu Nighttime Lights: Feb 24 - Mar 1, 2026",
       subtitle = "Before First Kona Low Storm of 2026") +
  coord_sf() +
  theme_void() +
  theme(plot.title = element_text(face = "bold", hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        strip.text = element_text(face = "bold"),
        legend.position = "none")
```

```{r}
#| echo: false

### Map: one panel per day, March 10-16, 2026
ggplot() +
  
  geom_spatraster(data = r_masked$storm1) +
  facet_wrap(~lyr, ncol = 2) +
  scale_fill_gradient2(low = "black",
                       mid = "yellow",
                       high = "red",
                       midpoint = 4.5,
                       na.value = "transparent") +
  
  labs(title = "Oahu Nighttime Lights: March 10-16, 2026",
       subtitle = "First Kona Low Storm of 2026 \n(Missing Mar 10-12)") +
  coord_sf() +
  theme_void() +
  theme(plot.title = element_text(face = "bold", hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        strip.text = element_text(face = "bold"),
        legend.position = "none")
```

```{r}
ggplot() +
  
  geom_spatraster(data = r_masked$storm2) +
  facet_wrap(~lyr, ncol = 3) +
  scale_fill_gradient2(low = "black",
                       mid = "yellow",
                       high = "red",
                       midpoint = 4.5,
                       na.value = "transparent") +
  
  labs(title = "Oahu Nighttime Lights: March 19-24, 2026",
       subtitle = "Second Kona Low Storm of 2026") +
  coord_sf() +
  theme_void() +
  theme(plot.title = element_text(face = "bold", hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        strip.text = element_text(face = "bold"),
        legend.position = "none")
```

```{r}
baseline_luminance <- terra::global(r_masked$baseline, "mean", na.rm = TRUE)[,1]
storm1_luminance <- terra::global(r_masked$storm1, "mean", na.rm = TRUE)[,1]
storm2_luminance <- terra::global(r_masked$storm2, "mean", na.rm = TRUE)[,1]

daily_dates <- c(all_dates$baseline, ymd(c("2026-03-13", "2026-03-14", "2026-03-15", "2026-03-16", "2026-03-19", "2026-03-20", "2026-03-21", "2026-03-22", "2026-03-23", "2026-03-24")))
daily_luminance <- c(baseline_luminance, storm1_luminance, storm2_luminance)

df_storm1 <- data.frame(
  Date = daily_dates,
  Luminance = daily_luminance,
  Period = c(rep("Baseline", length(all_dates$baseline)), rep("First Kona Low", 4), rep("Second Kona Low", 6))
)

ggplot(df_storm1, aes(x = Date, y = Luminance, fill = Period)) +
  geom_col() +
  scale_fill_manual(values = c("Baseline" = "gray70", "First Kona Low" = "steelblue", "Second Kona Low" = "maroon")) +
  labs(
    title = "Daily Overall Luminance: First Kona Low",
    x = "Date",
    y = "Mean Luminance",
    fill = "Period"
  ) +
  theme_minimal()
```

```{r}
ggplot() +
  geom_spatraster(data = absolute_storm1) +
  geom_spatvector(data = roi_sf, fill = NA, color = "black", linewidth = 0.2) +
  scale_fill_gradient2(
    low = "blue",
    mid = "white",
    high = "red",
    midpoint = 0,
    na.value = "transparent"
  ) +
  labs(
    title = "Absolute Change in Nighttime Lights",
    subtitle = "First Kona Low Storm vs. Baseline Period",
    fill = "NTL Change"
  ) +
  coord_sf() +
  theme_void() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  )
```

```{r}
ggplot() +
  geom_spatraster(data = absolute_storm2) +
  geom_spatvector(data = roi_sf, fill = NA, color = "black", linewidth = 0.2) +
  scale_fill_gradient2(
    low = "blue",
    mid = "white",
    high = "red",
    midpoint = 0,
    na.value = "transparent"
  ) +
  labs(
    title = "Absolute Change in Nighttime Lights",
    subtitle = "Second Kona Low Storm vs. Baseline Period",
    fill = "NTL Change"
  ) +
  coord_sf() +
  theme_void() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  )
```

```{r}
ggplot() +
  geom_spatraster(data = absolute_storm2_gap) +
  geom_spatvector(data = roi_sf, fill = NA, color = "black", linewidth = 0.2) +
  scale_fill_gradient2(
    low = "blue",
    mid = "white",
    high = "red",
    midpoint = 0,
    na.value = "transparent"
  ) +
  labs(
    title = "Absolute Change in Nighttime Lights",
    subtitle = "Second Kona Low Storm vs. Gap Day",
    fill = "NTL Change"
  ) +
  coord_sf() +
  theme_void() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  )
```

```{r}
ggplot() +
  geom_spatraster(data = percent_storm1) +
  geom_spatvector(data = roi_sf, fill = NA, color = "black", linewidth = 0.2) +
  scale_fill_gradient2(
    low = "blue",
    mid = "white",
    high = "red",
    midpoint = 0,
    na.value = "transparent"
  ) +
  labs(
    title = "Percent Change in Nighttime Lights",
    subtitle = "First Kona Low Storm vs. Baseline Period",
    fill = "Percent Change"
  ) +
  coord_sf() +
  theme_void() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  )
```

```{r}
ggplot() +
  geom_spatraster(data = percent_storm2) +
  geom_spatvector(data = roi_sf, fill = NA, color = "black", linewidth = 0.2) +
  scale_fill_gradient2(
    low = "blue",
    mid = "white",
    high = "red",
    midpoint = 0,
    na.value = "transparent"
  ) +
  labs(
    title = "Percent Change in Nighttime Lights",
    subtitle = "Second Kona Low Storm vs. Baseline Period",
    fill = "Percent Change"
  ) +
  coord_sf() +
  theme_void() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  )

```

```{r}
ggplot() +
  geom_spatraster(data = percent_storm2_gap) +
  geom_spatvector(data = roi_sf, fill = NA, color = "black", linewidth = 0.2) +
  scale_fill_gradient2(
    low = "blue",
    mid = "white",
    high = "red",
    midpoint = 0,
    na.value = "transparent"
  ) +
  labs(
    title = "Percent Change in Nighttime Lights",
    subtitle = "Second Kona Low Storm vs. Gap Day",
    fill = "Percent Change"
  ) +
  coord_sf() +
  theme_void() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  )
```
