---
title: "Isochrones"
format: html
editor: visual
---

An **isochrone** is a line or polygon on a map that connects all points reachable from a specific starting location within a set amount of time

[Inspiration from Kyle Walker](https://walker-data.com/posts/accessibility-surface/)

### Libraries

```{r}
library(mapboxapi)
library(leaflet)
library(mapview)
library(here)
library(sf)
library(tidyverse)
```

### Create Isochrone

```{r}
isos <- mb_isochrone(
  location = "One Bowerman Dr, Beaverton, OR 97005",
  profile = "driving",
  time = 1:45
)
```

```{r}
mapview(isos, zcol = "time")
```

```{r}
library(sf)
library(fasterize)

isos_proj <- st_transform(isos, 32618)

template <- raster(isos_proj, resolution = 100)

iso_surface <- fasterize(isos_proj, template, field = "time", fun = "min")

iso_surface
pal <- colorNumeric("plasma", isos$time, na.color = "transparent")

nike_map <- leaflet() %>%
  addMapboxTiles(style_id = "light-v9",
                 username = "mapbox",
                 scaling_factor = "0.5x") %>%
  addRasterImage(iso_surface, colors = pal, opacity = 0.5) %>%
  addLegend(values = isos$time, pal = pal,
            title = "Drive-time from<br>Nike HQ")

nike_map
```

```{r}
nike_icon <- makeIcon(
  iconUrl = "https://nike.com/favicon.ico",
  iconWidth = 25,
  iconHeight = 25
)

nike_hq <- mb_geocode("One Bowerman Dr, Beaverton, OR 97005")

nike_map %>%
  addMarkers(lng = nike_hq[1], lat = nike_hq[2], icon = nike_icon)
```

### Your turn - Recreate for Chaminade

```{r}
chaminade_isos <- mb_isochrone(
  location = "3140 Waialae Ave, Honolulu, HI 96816",
  profile = "driving",
  time = 1:30
)
```

```{r}
mapview(chaminade_isos, zcol = "time")
```

```{r}
chaminade_proj <- st_transform(chaminade_isos, 32618)

template <- raster(chaminade_proj, resolution = 100)

chaminade_surface <- fasterize(chaminade_proj, template, field = "time", fun = "min")

chaminade_surface
pal <- colorNumeric("plasma", chaminade_isos$time, na.color = "transparent")

chaminade_map <- leaflet() %>%
  addMapboxTiles(style_id = "light-v9",
                 username = "mapbox",
                 scaling_factor = "0.5x") %>%
  addRasterImage(chaminade_surface, colors = pal, opacity = 0.5) %>%
  addLegend(values = chaminade_isos$time, pal = pal,
            title = "Drive-time from<br>Chaminade University")

chamiande_icon <- makeIcon(
  iconUrl = "https://queens.org/favicon.ico",
  iconWidth = 25,
  iconHeight = 25
)

chamiande <- mb_geocode("3140 Waialae Ave, Honolulu, HI 96816")

chaminade_map |> 
  addMarkers(lng = chamiande[1], lat = chamiande[2], icon = chamiande_icon)
```

### Lets go further

Question: Can we make an accessibility surfaces map for trauma centers, ER's, hospitals, fire stations etc. Maybe combine all of them into one shiny app. Let's strategise and design...

**App Design:**

- Fire stations: mb_isochrone works well
- Ambulance:
- Shelters:
  - Can't go base off travel time for island-wide emergency situations due to possible increase in traffic
- **Flood risk zones to a flood free zone (Ashley & Faith H.)**
  - Using polygons to map high-risk flood zones and low-risk flood zones
    - [Tsunami Evacuation Zones](https://geoportal.hawaii.gov/datasets/HiStateGIS::tsunami-evacuation-zones/explore?location=21.410979%2C-158.148541%2C13)
    - [Extreme Tsunami Evacuation Zones](https://geoportal.hawaii.gov/datasets/extreme-tsunami-evacuation-zones/explore?location=21.410528%2C-158.148604%2C13)
    - [Oahu addresses with geometry](https://honolulu-cchnl.opendata.arcgis.com/datasets/address-points-1/explore?location=21.481900%2C-157.964450%2C10&showTable=true)
  - depart_at
  - Could reference the date: July 29, 2025 \@ 2:45 PM HST — This was when statewide tsunami warning was announced
    - Should be specified as an ISO 8601 date/time, e.g. `"2025-07-29T14:45"`
  - Think about who this would be for — I think it would be for individuals who live in high risk tsunami zones
- Hospitals, trauma centers, etc.
- Police stations

### Read in Tsunami Zones Data

```{r}
tsunami_zones <- st_read(here("data/Tsunami_Evacuation_Zones.geojson"))
extreme_tsunami_zones <- st_read(here("data/Extreme_Tsunami_Evacuation_Zones.geojson"))
oahu_adress <- st_read(here("data/oahu_address.geojson"))
```

### Filter Polygons to Oahu

```{r}
oahu_tsunami_zones <- tsunami_zones |> 
  filter(island == "OAHU")

oahu_extreme_tsunami_zones <- extreme_tsunami_zones |> 
  filter(island == "OAHU")


```

```{r}
mapview(oahu_tsunami_zones)
```

```{r}
mapview(oahu_extreme_tsunami_zones)
```

With the address data from oahu addresses, can we create a new dataframe with only the addresses that have the geometries in the tsunami zones from the tsunami zone geometries and the extreme tsunami zone geometries?

Should we combine tsunami and extreme tsunami into 1 polygon

Yes, but we keep the distinction . one isochrme, but distict boundaries

The idea is that we're mapping the distance from x address on a given day (tsunami day) and asking how far given this time of this day, how far am I from my address to safety? the overlay of the tsunami zones is what gives context to our design

```{r}
head(oahu_adress)
```
