---
title: "Isochrones"
format: html
editor: visual
---

An **isochrone** is a line or polygon on a map that connects all points reachable from a specific starting location within a set amount of time

[Inspiration from Kyle Walker](https://walker-data.com/posts/accessibility-surface/)

### Libraries

```{r}
library(mapboxapi)
library(leaflet)
library(mapview)
library(here)
library(sf)
library(tidyverse)
library(mapgl)
library(tigris)
```

### Create Isochrone

```{r}
isos <- mb_isochrone(
  location = "One Bowerman Dr, Beaverton, OR 97005",
  profile = "driving",
  time = 1:45
)
```

```{r}
mapview(isos, zcol = "time")
```

```{r}
library(sf)
library(fasterize)

isos_proj <- st_transform(isos, 32618)

template <- raster(isos_proj, resolution = 100)

iso_surface <- fasterize(isos_proj, template, field = "time", fun = "min")

iso_surface
pal <- colorNumeric("plasma", isos$time, na.color = "transparent")

nike_map <- leaflet() %>%
  addMapboxTiles(style_id = "light-v9",
                 username = "mapbox",
                 scaling_factor = "0.5x") %>%
  addRasterImage(iso_surface, colors = pal, opacity = 0.5) %>%
  addLegend(values = isos$time, pal = pal,
            title = "Drive-time from<br>Nike HQ")

nike_map
```

```{r}
nike_icon <- makeIcon(
  iconUrl = "https://nike.com/favicon.ico",
  iconWidth = 25,
  iconHeight = 25
)

nike_hq <- mb_geocode("One Bowerman Dr, Beaverton, OR 97005")

nike_map %>%
  addMarkers(lng = nike_hq[1], lat = nike_hq[2], icon = nike_icon)
```

### Your turn - Recreate for Chaminade

```{r}
chaminade_isos <- mb_isochrone(
  location = "3140 Waialae Ave, Honolulu, HI 96816",
  profile = "driving",
  time = 1:30
)
```

```{r}
mapview(chaminade_isos, zcol = "time")
```

```{r}
chaminade_proj <- st_transform(chaminade_isos, 32618)

template <- raster(chaminade_proj, resolution = 100)

chaminade_surface <- fasterize(chaminade_proj, template, field = "time", fun = "min")

chaminade_surface
pal <- colorNumeric("plasma", chaminade_isos$time, na.color = "transparent")

chaminade_map <- leaflet() %>%
  addMapboxTiles(style_id = "light-v9",
                 username = "mapbox",
                 scaling_factor = "0.5x") %>%
  addRasterImage(chaminade_surface, colors = pal, opacity = 0.5) %>%
  addLegend(values = chaminade_isos$time, pal = pal,
            title = "Drive-time from<br>Chaminade University")

chamiande_icon <- makeIcon(
  iconUrl = "https://queens.org/favicon.ico",
  iconWidth = 25,
  iconHeight = 25
)

chamiande <- mb_geocode("3140 Waialae Ave, Honolulu, HI 96816")

chaminade_map |> 
  addMarkers(lng = chamiande[1], lat = chamiande[2], icon = chamiande_icon)
```

### Lets go further

Question: Can we make an accessibility surfaces map for trauma centers, ER's, hospitals, fire stations etc. Maybe combine all of them into one shiny app. Let's strategise and design...

**App Design:**

- Fire stations: mb_isochrone works well
- Ambulance:
- Shelters:
  - Can't go base off travel time for island-wide emergency situations due to possible increase in traffic
- **Flood risk zones to a flood free zone (Ashley & Faith H.)**
  - Using polygons to map high-risk flood zones and low-risk flood zones
    - [Tsunami Evacuation Zones](https://geoportal.hawaii.gov/datasets/HiStateGIS::tsunami-evacuation-zones/explore?location=21.410979%2C-158.148541%2C13)
    - [Extreme Tsunami Evacuation Zones](https://geoportal.hawaii.gov/datasets/extreme-tsunami-evacuation-zones/explore?location=21.410528%2C-158.148604%2C13)
    - [Oahu addresses with geometry](https://honolulu-cchnl.opendata.arcgis.com/datasets/address-points-1/explore?location=21.481900%2C-157.964450%2C10&showTable=true)
  - depart_at
  - Could reference the date: July 29, 2025 \@ 2:45 PM HST — This was when statewide tsunami warning was announced
    - Should be specified as an ISO 8601 date/time, e.g. `"2025-07-29T14:45"`
  - Think about who this would be for — I think it would be for individuals who live in high risk tsunami zones
- Hospitals, trauma centers, etc.
- Police stations

### Read in Tsunami Zones Data

```{r}
tsunami_zones <- st_read(here("data/Tsunami_Evacuation_Zones.geojson"))
extreme_tsunami_zones <- st_read(here("data/Extreme_Tsunami_Evacuation_Zones.geojson"))
oahu_adress <- st_read(here("data/oahu_address.geojson"))
```

### Filter Polygons to Oahu

```{r}
oahu_tsunami_zones <- tsunami_zones |> 
  filter(island == "OAHU")

oahu_extreme_tsunami_zones <- extreme_tsunami_zones |> 
  filter(island == "OAHU")


```

```{r}
mapview(oahu_tsunami_zones)
```

```{r}
mapview(oahu_extreme_tsunami_zones)
```

With the address data from oahu addresses, can we create a new dataframe with only the addresses that have the geometries in the tsunami zones from the tsunami zone geometries and the extreme tsunami zone geometries?

Should we combine tsunami and extreme tsunami into 1 polygon

Yes, but we keep the distinction . one isochrme, but distict boundaries

The idea is that we're mapping the distance from x address on a given day (tsunami day) and asking how far given this time of this day, how far am I from my address to safety? the overlay of the tsunami zones is what gives context to our design

### Slimming address data

```{r}
address1 <- oahu_adress |>
  select(address_id, full_number, streetname, city, zip) |>
  filter(!st_is_empty(geometry)) 

address1
nrow(address1)
```

### preparing tsunami layers

from looking at the zones earlier I saw that in the extreme tsunami zone instead of being 1 polygon it was split up. so I merge it into one using st_union so when I check my addresses against the tsunami layer, instead of checking it against 20+ polygons it just checks against 1.

```{r}
tsunami1 <- oahu_tsunami_zones |>
  st_make_valid() |>
  summarise(geometry = st_union(geometry))

extreme1 <- oahu_extreme_tsunami_zones |>
  st_make_valid() |>
  summarise(geometry = st_union(geometry))

tsunami1
extreme1
```

I also keep the two zones seperate objects because I want to be able to tell them apart on my shinyapp. Their geometry also dont overlapp at all so I know no address can be inside both at the same time.

### labeling every address by zone 

```{r}
address_zones <- address1 |>
  mutate(
    in_tsunami = lengths(st_intersects(geometry, tsunami1)) > 0,
    in_extreme = lengths(st_intersects(geometry, extreme1)) > 0,#checking each adress against tsunami zone, lengths >0 turning each into true(isndie) or false(outside) zone for both extreme and regular zone
    zone = case_when(
      in_tsunami ~ "Tsunami Evacuation Zone",
      in_extreme ~ "Extreme Tsunami Evacuation Zone",
      TRUE       ~ "Outside evacuation zones"
    )
  ) #turns the two columns of true/false into one label, and true labels everything outside of evacuation zones

address_zones |>
  st_drop_geometry() |>
  count(in_tsunami, in_extreme, zone) #removes geometry column and turns sf into regular df, address zones still has all its points, for cleaness
```

time to start map stuff

### adding a search label and map the addresses

```{r}

# Readable address label for searching and tooltips
address_zones <- address_zones |>
  mutate(search_label = paste0(full_number, " ", streetname, ", ", city, " ", zip)) #this makes address parts one string, full_number keeps hyphen for search box searches later.

# Combine the two zone layers into one, with a label column
zones_map <- bind_rows(
  tsunami1 |> mutate(zone = "Tsunami Evacuation Zone"),
  extreme1 |> mutate(zone = "Extreme Tsunami Evacuation Zone")
) #stacks two zones into one two row layer for easier coloring

zone_vals <- c("Tsunami Evacuation Zone",
               "Extreme Tsunami Evacuation Zone",
               "Outside evacuation zones")
zone_cols <- c("#d7301f", "#fc8d59", "#9e9e9e") #stores labels and colors so legend polygons and points all match. red for standard zone, orange for extreme, and the gray for outside

mapboxgl(bounds = zones_map) |>
  add_fill_layer(
    id = "zones",
    source = zones_map,
    fill_color = match_expr("zone", values = zone_vals[1:2], stops = zone_cols[1:2]),
    fill_opacity = 0.35 #changes opacity so roads are visible
  ) |>
  add_circle_layer(
    id = "addresses",
    source = address_zones,
    circle_color = match_expr("zone", values = zone_vals, stops = zone_cols),
    circle_radius = 1.5,
    tooltip = "search_label" #draws addresses as small dots
  ) |>
  add_legend(
    "Evacuation zone",
    values = zone_vals,
    colors = zone_cols,
    type = "categorical"
  )
```

### testing traffic isochrome for one address 

```{r}
set.seed(422)
test_addr <- address_zones |>
  filter(zone == "Tsunami Evacuation Zone") |>
  slice_sample(n = 1) #choosing random test address

test_addr$search_label


tsunami_day <- "2026-10-13T15:00" #chosen time based on warning

test_iso <- mb_isochrone(
  test_addr,
  time = c(5, 10, 15, 20),
  profile = "driving-traffic",
  depart_at = tsunami_day 
)

test_iso
```

### rings over zones

```{r}
ring_cols <- c("#1a9850", "#91cf60", "#fee08b", "#4575b4")

mapboxgl(bounds = test_iso) |>
  add_fill_layer(
    id = "zones",
    source = zones_map,
    fill_color = match_expr("zone", values = zone_vals[1:2], stops = zone_cols[1:2]),
    fill_opacity = 0.35
  ) |>
  add_line_layer(
    id = "rings",
    source = test_iso,
    line_color = match_expr("time", values = c(5, 10, 15, 20), stops = ring_cols),
    line_width = 2
  ) |>
  add_markers(data = test_addr) |>
  add_legend(
    "Drive time (minutes)",
    values = c("5", "10", "15", "20"),
    colors = ring_cols,
    type = "categorical"
  )
```

```{r}
# Both zones together: everywhere unsafe in an extreme warning
unsafe_all <- st_union(c(st_geometry(tsunami1), st_geometry(extreme1)))

# How much of a ring (in m²) lies outside an unsafe area
area_outside <- function(rings, unsafe) {
  sapply(seq_len(nrow(rings)), function(i) {
    left <- st_difference(st_geometry(rings)[i], st_geometry(unsafe))
    if (length(left) == 0) 0 else sum(as.numeric(st_area(left)))
  })
}

test_iso <- arrange(test_iso, time)

test_iso$m2_outside_tsunami <- area_outside(test_iso, tsunami1)
test_iso$m2_outside_all     <- area_outside(test_iso, unsafe_all)

st_drop_geometry(test_iso)
```

```{r}
safe_part_5 <- st_difference(st_geometry(test_iso)[1], unsafe_all)

mapboxgl(bounds = test_iso[1, ]) |>
  add_fill_layer(id = "zones", source = zones_map,
                 fill_color = match_expr("zone", values = zone_vals[1:2], stops = zone_cols[1:2]),
                 fill_opacity = 0.35) |>
  add_fill_layer(id = "safe5", source = st_sf(geometry = safe_part_5),
                 fill_color = "#1a9850", fill_opacity = 0.6) |>
  add_markers(data = test_addr)
```

fixing spill over

```{r}
# Oahu's land area, clipped to the coastline
oahu_land <- counties(state = "HI", cb = TRUE) |>
  filter(NAME == "Honolulu") |>
  st_transform(4326) |>
  st_make_valid()

# Safe ground for each warning type
safe_standard <- st_difference(st_geometry(oahu_land), st_geometry(tsunami1))
safe_extreme  <- st_difference(st_geometry(oahu_land), unsafe_all)

# How much of a ring (in m²) overlaps safe ground
area_safe <- function(rings, safe) {
  sapply(seq_len(nrow(rings)), function(i) {
    hit <- st_intersection(st_geometry(rings)[i], safe)
    if (length(hit) == 0) 0 else sum(as.numeric(st_area(hit)))
  })
}

test_iso$m2_safe_standard <- area_safe(test_iso, safe_standard)
test_iso$m2_safe_extreme  <- area_safe(test_iso, safe_extreme)

# Compare the old and new measurements
test_iso |>
  st_drop_geometry() |>
  select(time, m2_outside_tsunami, m2_safe_standard, m2_outside_all, m2_safe_extreme)
```

```{r}
# Smallest ring time that reaches meaningful safe land
first_safe <- function(rings, safe, min_area = 5000) {
  rings <- arrange(rings, time)
  m2 <- area_safe(rings, safe)
  hits <- rings$time[m2 > min_area]
  if (length(hits) > 0) min(hits) else NA_real_
}

minutes_to_safety <- function(addr, depart_at = tsunami_day) {
  
  # Already safe: no API call needed
  if (addr$zone == "Outside evacuation zones") {
    return(tibble(exit_standard = 0, exit_extreme = 0))
  }
  
  get_rings <- function(times) {
    mb_isochrone(addr, time = times, profile = "driving-traffic", depart_at = depart_at)
  }
  
  # Main call: 5, 10, 15, 20 minute rings
  rings <- get_rings(c(5, 10, 15, 20))
  
  exit_standard <- if (addr$zone == "Tsunami Evacuation Zone") first_safe(rings, safe_standard) else 0
  exit_extreme  <- first_safe(rings, safe_extreme)
  
  # Sharpen the answer if safety is reached within 5 minutes
  if (isTRUE(exit_standard == 5) || isTRUE(exit_extreme == 5)) {
    fine <- get_rings(c(1, 2, 3, 4))
    if (isTRUE(exit_standard == 5)) exit_standard <- coalesce(first_safe(fine, safe_standard), 5)
    if (isTRUE(exit_extreme == 5))  exit_extreme  <- coalesce(first_safe(fine, safe_extreme), 5)
  }
  
  # Extend the search if safety isn't reached within 20 minutes
  if (is.na(exit_standard) || is.na(exit_extreme)) {
    long <- get_rings(c(30, 40, 50, 60))
    if (is.na(exit_standard)) exit_standard <- first_safe(long, safe_standard)
    if (is.na(exit_extreme))  exit_extreme  <- first_safe(long, safe_extreme)
  }
  
  tibble(exit_standard = exit_standard, exit_extreme = exit_extreme)
}

minutes_to_safety(test_addr)
```

this model uses normal traffic instead of the evacuation time. it also measures time to "safe" land, not a destination. this isnt accruate. people dont travel to a spot of land, they travel to a destination.

### testing zones

```{r}
set.seed(422)

test_extreme <- address_zones |>
  filter(zone == "Extreme Tsunami Evacuation Zone") |>
  slice_sample(n = 1)

test_outside <- address_zones |>
  filter(zone == "Outside evacuation zones") |>
  slice_sample(n = 1)

test_extreme$search_label
minutes_to_safety(test_extreme)

test_outside$search_label
minutes_to_safety(test_outside)
```
