---
title: "GIS for Natural Diaster Response"
format: html
editor: visual
---

### Welcome to our first DS 422 notebook

With the recent power outages from Hurricane Lala, let's explore how we can apply GIS skills to natural disaster management.

#### Data

We will start with the [Eagle-I Power Outage Dataset for 2025](https://www.osti.gov/biblio/3012826) from Oak Ridge National Laboratory.

#### Libraries

```{r}
library(tidyverse)
library(here)
library(sf)
library(tidycensus)
library(arrow)
library(tigris)
library(sf)
library(ggiraph)
library(plotly)
```

I downloaded the original dataset from the link above, filtered to only Hawaii outages, and converted the csv to a feather file to shrink the size for ease of sharing and loading

```{r}
eagle_i_hawaii <- read_feather(here("data/hawaii_outages_2025.feather"))
```

#### Your Turn - Exploratory Data Analysis

Take 15 minutes to explore trends in this data, think summary stats, plots, etc..

```{r}

```

```{r}

```

#### Outages by county

##### First filter out kalawao

```{r}
eagle_i_hawaii <- eagle_i_hawaii |> 
  filter(county!= "Kalawao")
```

##### Next, we can count outages by county

```{r}
customers_out_by_county <- eagle_i_hawaii |> 
  group_by(county) |> 
  summarize(total_customers_out = sum(customers_out, na.rm = TRUE))
```

##### Your turn - Lets make a bar chart with ggplot

```{r}
ggplot(data= customers_out_by_county, aes(y=county, x=total_customers_out)) +geom_col()

#ggplot(customers_out_by_county_with_pop, aes(y = county, x = outages_per_1000)) +
 # geom_col(fill = "firebrick") +
  #labs(
   # title = "Power Outages per 1,000 Residents by County",
    ##x = "Outages per 1,000 residents",
    #y = NULL
  #) +
  #theme_minimal()
```

##### What if we normalize?

```{r}
hi_pop <- get_acs(
  geography = "county",
  state = "HI",
  variables = "B01003_001",  # total population
  year = 2022
) 
```

##### Joining our data

```{r}
hi_pop <- hi_pop |>
  mutate(county = gsub(" County, Hawaii", "", NAME))

hi_pop <- hi_pop |>
  select(county, population = estimate)

customers_out_by_county_with_pop <- customers_out_by_county |>
  left_join(hi_pop, by = "county")
```

##### Normalizing

```{r}
customers_out_by_county_with_pop <- customers_out_by_county_with_pop |>
  mutate(outages_per_1000 = (total_customers_out / population) * 1000)
```

##### Now Plot

```{r}
ggplot(customers_out_by_county_with_pop, aes(y = county, x = outages_per_1000)) +
  geom_col(fill = "firebrick") +
  labs(
    title = "Power Outages per 1,000 Residents by County",
    x = "Outages per 1,000 residents",
    y = NULL
  ) +
  theme_minimal()
```

#### Outages over time

```{r}
outages_over_time <- ggplot(data = eagle_i_hawaii, aes (x = run_start_time, y = customers_out)) +
  geom_line(linewidth = 0.2) +
  theme_minimal()
outages_over_time
```

```{r}
ggplotly(outages_over_time)
```

##### Color by county

```{r}
county_outages_over_time <- ggplot(data = eagle_i_hawaii, aes (x = run_start_time, y = customers_out, color = county)) +
  geom_line(linewidth = 0.2) +
  theme_minimal()
county_outages_over_time
```

```{r}
ggplotly(county_outages_over_time)
```

##### Facet by county

```{r}
ggplot(data = eagle_i_hawaii, aes (x = run_start_time, y = customers_out)) +
  geom_line(linewidth = 0.2) +
  facet_wrap(~county, nrow = 4, scales = "free") +
  theme_bw()
```

#### Map Outages

```{r}
library(tigris)
library(sf)
library(ggplot2)
library(dplyr)
library(patchwork)

# Get Hawaii county boundaries
hi_counties_sf <- counties(state = "HI", cb = TRUE, year = 2022) |>
  mutate(county = gsub(" County", "", NAME)) |>
  filter(county != "Kalawao") |>
  left_join(customers_out_by_county_with_pop, by = "county") |> 
  st_crop(xmin = -160.5, xmax = -154.5, ymin = 18.5, ymax = 22.5)

```

```{r}
ggplot(hi_counties_sf) +
  geom_sf(aes(fill = outages_per_1000)) +
  scale_fill_viridis_c(option = "inferno", name = "Outages\nper 1,000") +
  labs(title = "Power Outages per 1,000 Residents by County") +
  theme_void()
```

##### Bonus

```{r}
p1 <- ggplot(hi_counties_sf, aes(
  x = reorder(county, outages_per_1000),
  y = outages_per_1000,
  tooltip = paste0(county, ": ", round(outages_per_1000, 1), " per 1,000"),
  data_id = county,
  fill = outages_per_1000
)) +
  geom_col_interactive() +
  coord_flip() +
  scale_fill_viridis_c(option = "inferno") +
  labs(x = NULL, y = "Outages per 1,000 residents", fill = NULL) +
  theme_minimal() +
  theme(legend.position = "none")

# ------------------------------------------------------------
# 3. Choropleth map (interactive, severity-colored)
# ------------------------------------------------------------

p2 <- ggplot(hi_counties_sf) +
  geom_sf_interactive(
    aes(fill = outages_per_1000,
        tooltip = paste0(county, ": ", round(outages_per_1000, 1), " per 1,000"),
        data_id = county),
    color = "white"
  ) +
  scale_fill_viridis_c(option = "inferno", name = "Outages\nper 1,000") +
  coord_sf() +
  theme_void()

# ------------------------------------------------------------
# 4. Combine and make interactive
# ------------------------------------------------------------

combined_plot <- p1 + p2 + plot_layout(widths = c(1, 1.2))

interactive_plot <- girafe(ggobj = combined_plot, width_svg = 10, height_svg = 5)
interactive_plot <- girafe_options(
  interactive_plot,
  opts_hover(css = "stroke:black;stroke-width:2px;"),
  opts_hover_inv(css = "opacity:0.3;")
)

interactive_plot
```

#### Whats Next?

This seems like a good start to understanding power outages, but is fairly limited with the county level granularity of the data. Another option to measure power outages is utilizing the [blackmarbler](https://worldbank.github.io/blackmarbler/index.html) package.

**BlackMarbleR** is a R package that provides a simple way to use nighttime lights data from NASA’s Black Marble. [Black Marble](https://blackmarble.gsfc.nasa.gov/) is a [NASA Earth Science Data Systems (ESDS)](https://www.earthdata.nasa.gov/) project that provides a product suite of daily, monthly and yearly global [nighttime lights](https://www.earthdata.nasa.gov/topics/human-dimensions/nighttime-lights). This package automates the process of downloading all relevant tiles from the [NASA LAADS DAAC](https://ladsweb.modaps.eosdis.nasa.gov/) to cover a region of interest, converting and mosaicing the raw files (in HDF5 format) to georeferenced rasters.

#### Your Turn

[Follow the steps here](https://worldbank.github.io/blackmarbler/#installation) to generate the first 2 visualizations of nighttime lights in Ghana (map and bar chart). Stop at "Workflow to Update Data". We will tweak this analysis to measure power outages in Honolulu county pre/during/post severe weather events with known outages from the eagle-i dataset

```{r}
#libraries
#### Setup
# Load packages
library(blackmarbler)
library(geodata)
library(sf)
library(terra)
library(ggplot2)
library(tidyterra)
library(lubridate)

#### Define NASA bearer token
bearer <- "eyJ0eXAiOiJKV1QiLCJvcmlnaW4iOiJFYXJ0aGRhdGEgTG9naW4iLCJzaWciOiJlZGxqd3RwdWJrZXlfb3BzIiwiYWxnIjoiUlMyNTYifQ.eyJ0eXBlIjoiVXNlciIsInVpZCI6ImZoYXJkaWUiLCJleHAiOjE3OTI0NTYwNzksImlhdCI6MTc4NzI3MjA3OSwiaXNzIjoiaHR0cHM6Ly91cnMuZWFydGhkYXRhLm5hc2EuZ292IiwiaWRlbnRpdHlfcHJvdmlkZXIiOiJlZGxfb3BzIiwiYWNyIjoiZWRsIiwiYXNzdXJhbmNlX2xldmVsIjozfQ.3Lrj1YrkkqpbpmOhfkSlzLe7XHfQ3wULnJmFXRF5YyMgBBM2RYFUkZbrlmvHYqa-w0BbSRXs4lKvNh2wOBd_2YRwBXDcKmcNrqx8PivdFrDvCHIe5OBrnE3E_WvbAq3ogFBvZIPI61hg0R5NRhiCvyBrkGPWM89G5Ll9kRTZ9TdAbgf3Htudl_xbo4-wziTq4H0_MbLKThPEVf1Tl9ws5CFzia7CAGdMqpufXpAkbCEGG_UWA8ODTBTvPVSyxwW6G-42tuUkY1d8JcHBjtQvfDKiFo9miAdIB_GEIYduNujHHBokmsW9onfsHMr2RQCJU4y7C3ZnlEBRwDEZQIoJTg"

### ROI
# Define region of interest (roi). The roi must be (1) an sf polygon and (2)
# in the WGS84 (epsg:4326) coordinate reference system. Here, we use the
# getData function to load a polygon of Ghana
roi_sf <- gadm(country = "GHA", level=1, path = tempdir()) 
```

```{r}
### Daily data: raster for February 5, 2021
r_20210205 <- bm_raster(roi_sf = roi_sf,
                        product_id = "VNP46A2",
                        date = "2021-02-05",
                        bearer = bearer)

### Monthly data: raster for October 2021
r_202110 <- bm_raster(roi_sf = roi_sf,
                      product_id = "VNP46A3",
                      date = "2021-10-01", # The day is ignored
                      bearer = bearer)

### Annual data: raster for 2021
r_2021 <- bm_raster(roi_sf = roi_sf,
                    product_id = "VNP46A4",
                    date = 2021,
                    bearer = bearer)
```

```{r}
#### Daily data in March 2021
r_daily <- bm_raster(roi_sf = roi_sf,
                     product_id = "VNP46A2",
                     date = seq.Date(from = ymd("2021-03-01"), to = ymd("2021-03-31"), by = "day"),
                     bearer = bearer)

#### Monthly aggregated data in 2021 and 2022
r_monthly <- bm_raster(roi_sf = roi_sf,
                       product_id = "VNP46A3",
                       date = seq.Date(from = ymd("2021-01-01"), to = ymd("2022-12-01"), by = "month"),
                       bearer = bearer)

#### Yearly aggregated data in 2012 and 2021
r_annual <- bm_raster(roi_sf = roi_sf,
                      product_id = "VNP46A4",
                      date = 2012:2021,
                      bearer = bearer)
```

```{r}
#### Make raster
r <- bm_raster(roi_sf = roi_sf,
               product_id = "VNP46A3",
               date = "2021-10-01",
               bearer = bearer)

#### Prep data
r <- r |> terra::mask(roi_sf)

## Distribution is skewed, so log
r[] <- log(r[] + 1)

##### Map
ggplot() +
  geom_spatraster(data = r) +
  scale_fill_gradient2(low = "black",
                       mid = "yellow",
                       high = "red",
                       midpoint = 4.5,
                       na.value = "transparent") +
  labs(title = "Nighttime Lights: October 2021") +
  coord_sf() +
  theme_void() +
  theme(plot.title = element_text(face = "bold", hjust = 0.5),
        legend.position = "none")
```

```{r}
#### Extract annual data
ntl_df <- bm_extract(roi_sf = roi_sf,
                     product_id = "VNP46A4",
                     date = 2012:2022,
                     bearer = bearer)

#### Trends over time
ntl_df |>
  ggplot() +
  geom_col(aes(x = date,
               y = ntl_mean),
           fill = "darkorange") +
  facet_wrap(~NAME_1) +
  labs(x = NULL,
       y = "NTL Luminosity",
       title = "Ghana Admin Level 1: Annual Average Nighttime Lights") +
  scale_x_continuous(labels = seq(2012, 2022, 4),
                     breaks = seq(2012, 2022, 4)) +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold"))
```

```{r}
 
```
